---
title: "grayleafspotr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{grayleafspotr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(grayleafspotr)
```

## 1. Load the package

Open `grayleafspotr/grayleafspotr.Rproj` in RStudio for the source workflow,
or install the package and load it with `library(grayleafspotr)`.

## 2. Confirm the example data

```{r}
example_dir <- system.file("extdata", "example", package = "grayleafspotr")
list.files(example_dir)
```

## 3. Load saved example results

```{r}
example_run <- example_grayleafspot_results()
example_run$run
example_run$results
```

## 4. Generate template plots

```{r}
plot_colony_expansion(example_run)
plot_growth_roughness(example_run)
plot_stress_remodeling(example_run)
plot_texture_organization(example_run)
plot_shape_vs_stress(example_run)
plot_feature_heatmap(example_run)
plot_radial_profile(example_run)
```

## 5. Run the Python analysis pipeline on real images

### One-time Python setup

Add the interpreter path to `~/.Rprofile` so it loads automatically
every time R starts:

```r
file.edit("~/.Rprofile")
# Add this line (adjust path to your rvenv_arm_311 location):
# Sys.setenv(GRAYLEAFSPOTR_PYTHON = "/path/to/rvenv_arm_311/bin/python")
```

Or set it for the current session only:

```r
Sys.setenv(GRAYLEAFSPOTR_PYTHON = "/path/to/rvenv_arm_311/bin/python")
```

### Run with `grayleafspot_run()` — the primary entry point

`grayleafspot_run()` finds the installed package Python directory automatically,
creates the output folder if needed, and returns parsed JSON results directly.

```{r, eval=FALSE}
res <- grayleafspot_run(
  input_dir  = system.file("extdata", "testdata", "06FEB", package = "grayleafspotr"),
  output_dir = file.path(tempdir(), "gls_run"),
  run_name   = "trial_01"
)

# Access results
res$run       # manifest metadata
res$results   # per-image data frame
```

### Full-featured alternative: `grayleafspot_analyze()`

For tidy S3 objects with built-in plotting helpers, use `grayleafspot_analyze()`:

```{r, eval=FALSE}
real_run <- grayleafspot_analyze(
  input_dir  = system.file("extdata", "testdata", "06FEB", package = "grayleafspotr"),
  output_dir = tempdir(),
  save_outputs = FALSE,
  verbose      = FALSE,
  engine_model = "localunet"
)
plot_colony_expansion(real_run)
```

## 6. Convert to tidy data

```{r}
growth_data <- as_grayleafspot_growth_data(example_run)
growth_data
```

## 7. Custom plotting

```{r}
plot_colony_expansion(example_run) +
  ggplot2::labs(title = "Custom colony expansion view")
```
