---
title: "grayleafspotr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{grayleafspotr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(grayleafspotr)
```

## 1. Load the package

Open `R-Grayleafspot/grayleafspotr.Rproj` in RStudio for the source workflow,
or install the package and load it with `library(grayleafspotr)`.

## 2. Confirm the example data

```{r}
example_dir <- system.file("extdata", "example", package = "grayleafspotr")
list.files(example_dir)
```

## 3. Load saved example results

```{r}
example_run <- example_grayleafspot_results()
example_run$run
example_run$results
```

## 4. Generate template plots

```{r}
plot_colony_expansion(example_run)
plot_growth_roughness(example_run)
plot_stress_remodeling(example_run)
plot_texture_organization(example_run)
plot_shape_vs_stress(example_run)
plot_feature_heatmap(example_run)
plot_radial_profile(example_run)
```

## 5. Run the Python analysis pipeline on real images

Set `GRAYLEAFSPOTR_PYTHON` to the `rvenv_arm_311` interpreter before running:

```r
Sys.setenv(GRAYLEAFSPOTR_PYTHON = "/path/to/rvenv_arm_311/bin/python")
```

```{r, eval=FALSE}
if (grayleafspot_python_available(engine_model = "localunet")) {
  testdata_dir <- system.file("extdata", "testdata", "06FEB", package = "grayleafspotr")
  real_run <- grayleafspot_analyze(
    input_dir = testdata_dir,
    output_dir = tempdir(),
    save_outputs = FALSE,
    verbose = FALSE,
    engine_model = "localunet"
  )
  real_run$run
  real_run$results
} else {
  message("Python ML dependencies are not installed; reusing the example run.")
}
```

## 6. Convert to tidy data

```{r}
growth_data <- as_grayleafspot_growth_data(example_run)
growth_data
```

## 7. Custom plotting

```{r}
plot_colony_expansion(example_run) +
  ggplot2::labs(title = "Custom colony expansion view")
```
